---
title: "Untitled"
author: "Emmanouil Tranos"
date: "2022-09-09"
output: html_document
---

Street network analysis has been central to network science since its nascence: its mathematical foundation, graph theory, was born in the 18th century when Leonhard Euler presented his famous Seven Bridges of Königsberg problem.

Graphs can also have parallel (i.e., multiple) edges between the same two nodes.

network may be thought of as a real-world graph

A complex network is one with a nontrivial topology (the configuration and structure of its nodes and edges) -- that is, the topology is neither fully regular nor fully random. Most large real-world networks are complex (Newman, 2010). Of particular interest to this study are complex spatial networks -- that is, complex networks with nodes and/or edges embedded in space (O'Sullivan, 2014). A street network is an example of a complex spatial networkwith both nodes and edges embedded in space, as are railways, power grids, and water and sewage networks (Barthélemy, 2011).

street networks are approximately planar

Average street length

Node density

The eccentricity of a node is the maximum of the shortest-path weighted distances between it and each other node in the network

The diameter of a network is the maximum eccentricity of any node in the network and the radius of a network is theminimumeccentricity of any node in the network. The center of a network is the node or set of nodes whose eccentricity equals the radius and the periphery of a network is the node or set of nodeswhose eccentricity equals the diameter

node degree

clustering coefficient

The maximum betweenness centrality in a network specifies the proportion of shortest paths that pass through themost important node/edge. This is an indicator of resilience: networkswith a highmaximumbetweenness centrality aremore prone to failure or inefficiency should this single choke point fail.

Closeness centrality represents, for each node, the reciprocal of the sum of the distance from this node to all others in the network: that is, nodes rank as more central if they are on average closer to all other nodes.

Finally, PageRank -- the algorithm Google uses to rank web pages -- is a variant of network centrality, namely eigenvector centrality (Brin & Page, 1998). PageRank ranks nodes based on the structure of incoming links and the rank of the source node, andmay also be applied to street networks (Agryzkov, Oliver, Tortosa, & Vicent, 2012; Chin &Wen, 2015)

OpenStreetMap -- a collaborative mapping project that provides a free and publicly editable map of the world -- has emerged in recent years as a major player both for mapping and for acquiring spatial data (Corcoran, Mooney, & Bertolotto, 2013; Jokar Arsanjani, Zipf, Mooney, & Helbich, 2015). Inspired by Wikipedia's mass-collaboration model, the project started in 2004 and has grown to over two million users today. Its data quality is generally quite high (Barron, Neis, & Zipf, 2014; Girres & Touya, 2010; Haklay, 2010) and although data coverage varies worldwide, it is generally good when compared to corresponding estimates from the CIA World Factbook (Maron, 2015)

A city's development eras, design paradigms, underlying terrain, culture, and local economic conditions influence the pattern, topology, and grain of its street networks (Jackson 1985; Kostof 1991). These networks in turn structure the human interactions and transportation processes that run along them, forming an important pillar of city planners' quest for spatial order (Rose-Redwood and Bigon 2018). In particular, network orientation and geometry have played an outsized role in urban planning since its earliest days (Smith 2007). Measuring these network patterns can help researchers, planners, and community members understand local histories of urban design, transportation planning, and morphology; evaluate existing transportation system patterns and configurations; and explore new infrastructure proposals and alternatives. It also furthers the science of cities by providing a better understanding of urban patterns and how they correspond to evolutionary mechanisms, planning, and design.

The below are based on the very helpful vignettes of the [`sfnetworks`](https://cran.r-project.org/package=sfnetworks) and [``]()
packages. Please make sure to revisit these materials on your own time.

```{r setup, include=TRUE, results= 'hide', message=FALSE}
library(sfnetworks) #tidy geospatial networks
library(sf)         #simple features, spatial data
library(tidygraph)  #tidyverse's version of igraph
library(tidyverse)  #you should know this by now
library(igraph)     #as above
library(osmdata)    #access OSM data
library(dbscan)     #this is the dbscan clustering algorithm

knitr::opts_chunk$set(echo = TRUE)
```

In geospatial networks:

-  nodes always contain geographical coordinates (`sf` objects), 

-  edges can:

    - either contain geographical information (e.g. road networks)
    
    - or not, and instead just link two nodes (e.g. geospatial social networks)

The below code chunks create and manipulate a toy geospatial network.

```{r}
p1 = st_point(c(7, 51))
p2 = st_point(c(7, 52))
p3 = st_point(c(8, 52))
p4 = st_point(c(8, 51.5))

l1 = st_sfc(st_linestring(c(p1, p2)))
l2 = st_sfc(st_linestring(c(p1, p4, p3)))
l3 = st_sfc(st_linestring(c(p3, p2)))

edges = st_as_sf(c(l1, l2, l3), crs = 4326)
nodes = st_as_sf(c(st_sfc(p1), st_sfc(p2), st_sfc(p3)), crs = 4326)

edges$from = c(1, 1, 3)
edges$to = c(2, 3, 2)

net = sfnetwork(nodes, edges)

net

class(net)
```

Just like `igraph` we can refer to nodes using the `names` we can assign to them.

```{r}
nodes$name = c("city", "village", "farm")
edges$from = c("city", "city", "farm")
edges$to = c("village", "farm", "village")

edges

net = sfnetwork(nodes, edges, node_key = "name")
```

`net`'s edges contain geographical information. 
So, let's create an `other_net` whose edges do not contain geographical information.
This could have been a geospatial social network, for example. 
We know where each node is located in space and then the edges illustrate the existence 
of a connection between two nodes. 
To do so, we simply set the `st_geometry` equal to NULL.
The plots below will make these differences clear.

```{r}
st_geometry(edges) = NULL

other_net = sfnetwork(nodes, edges, edges_as_lines = TRUE)

plot(net, cex = 2, lwd = 2, main = "Original geometries")
plot(other_net, cex = 2, lwd = 2, main = "Geospatial social network example")
```
Let's increase the complexity. We can use the `roxel` data from the `sfnetworks`
package. Roxel is a small town just outise Munster in Germany.

The `as_sfnetwork()` function converts the spatial object to a network.
So, this is the spatial object and this is how it's geometry looks like.

```{r}
roxel

plot(st_geometry(roxel))
```

Now, we can convert it to a geospatial network and plot it.

```{r}
net <- as_sfnetwork(roxel)
plot(net)
```
Instead of `plot()` we can also use `ggplot2` functions. `autoplot()` does what it says.

```{r}
autoplot(net) + 
  ggtitle("Road network of Münster Roxel")
```

We have a network, so let's do some network analysis. 

First thing we need to do is to *activate* the network component we are interested 
in. In the below example we want to calculate the betweenness centrality of nodes 
and, therefore, we need to first activate the nodes.

So, the `net` object now has a `bc` column in the node attributes

```{r}
net = net %>%
  activate("nodes") %>%
  mutate(bc = centrality_betweenness())

net
```

We can now create a map of this spatial network, the size and the colour of its 
nodes are based on their betweenness centrality.

```{r}
ggplot() +
  geom_sf(data = st_as_sf(net, "edges"), col = "grey50") +
  geom_sf(data = st_as_sf(net, "nodes"), aes(col = bc), size = 3) + #, size = bc
  ggtitle("Betweenness centrality in Münster Roxel")
```

```{block, type='alert alert-warning'}
**Question**: Can you  interpret betweenness centrality in this context?
```

If we want to get rid of geometries, all we need to do is to remove the geometry.
As per base R, setting something equal to `NULL` means that we remove it.

```{r}
net %>%
  activate("edges") %>%
  st_set_geometry(NULL) %>%
  plot(draw_lines = T, main = "Edges without geometries")

net %>%
  activate("nodes") %>%
  st_set_geometry(NULL) %>%
  plot(vertex.color = "black", main = "Nodes without geometries")
```

OSM

```{r, eval=FALSE}

q <- opq(bbox = 'greater london uk') %>%
    add_osm_feature(key = 'highway', value = 'motorway')

x <- opq(bbox = 'greater london uk') %>%
    add_osm_feature(key = 'highway', value = 'motorway') %>%
    osmdata_sf()

x <- opq(bbox = 'greater london uk') %>%
    add_osm_feature(key = 'highway') %>%
    osmdata_sf()


plot(x$osm_points[1])

class(x)

class(osmdata_sf(q)$osm_lines)
class(osmdata_sf(x)$osm_lines)


q1 <- opq('Sevilla') %>%
    add_osm_feature(key = 'highway', value = 'cycleway')
cway_sev <- osmdata_sp(q1)
sp::plot(cway_sev$osm_lines)
```

To understand the `key` and `value` options, go to OSM's wiki [page](https://wiki.openstreetmap.org/wiki/Map_features#Highway)

```{r}
bristol_data <- opq(bbox = 'Bristol') %>%
  add_osm_feature(
    key = 'highway', 
    value = c("motorway", "trunk", "primary", "secondary", "tertiary", "residential", "unclassified", "motorway_link", "trunk_link", "primary_link", "secondary_link","tertiary_link", "residential_link")
) %>% 
  osmdata_sf(quiet = FALSE)

# extract the roads
bristol_roads <- st_geometry(bristol_data$osm_lines)
  
# build the sfnetwork object
bristol_sfn <- as_sfnetwork(bristol_roads, directed = FALSE, length_as_weight = TRUE)

plot(bristol_sfn)

ggplot() +
  geom_sf(data = st_as_sf(bristol_sfn, "edges"), col = "grey50") +
  geom_sf(data = st_as_sf(bristol_sfn, "nodes"), size = .01)
```

Network data wrangling. We can: 

1. remove multiple edges and loops. AKA simplifying the networks

2. create nodes where two lines intersect

3. remove nodes that are interior points in a line and define the shape of the 
line. AKA smoothing the network

```{r}
bristol_wrangled = bristol_sfn %>%
  activate("edges") %>%
  filter(!edge_is_multiple()) %>%
  filter(!edge_is_loop()) 

# size of original network
ecount(bristol_sfn) #number of edges
gorder(bristol_sfn) #number of nodes

# size of simplified network
ecount(bristol_wrangled)
gorder(bristol_wrangled)

bristol_wrangled <- convert(bristol_wrangled, to_spatial_subdivision)

# size of simplified network
ecount(bristol_wrangled)
gorder(bristol_wrangled)

bristol_wrangled <- convert(bristol_wrangled, to_spatial_smooth)

# size of simplified network
ecount(bristol_wrangled)
gorder(bristol_wrangled)
```
And plot again the network again.

```{r}
ggplot() +
  geom_sf(data = st_as_sf(bristol_wrangled, "edges"), col = "grey50") +
  geom_sf(data = st_as_sf(bristol_wrangled, "nodes"), size = .01)
```

```{r}
node_coords <- bristol_wrangled %>%
  activate("nodes") %>%
  st_coordinates()

clusters <- dbscan(node_coords, eps = 0.5, minPts = 1)$cluster

# Add the cluster information to the nodes of the network.
clustered <- bristol_wrangled %>%
  activate("nodes") %>%
  mutate(cls = clusters)

clustered <- clustered %>%
  mutate(cmp = group_components())

select(clustered, cls, cmp)

contracted = convert(
  clustered,
  to_spatial_contracted,
  cls, cmp,
  simplify = TRUE
)

ggplot() +
  geom_sf(data = st_as_sf(bristol_wrangled, "edges"), col = "grey50") +
  geom_sf(data = st_as_sf(bristol_wrangled, "nodes"), size = .01)
```


