---
title: "Network models in practice"
author: Emmanouil Tranos
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
editor: source
bibliography: references.bib
---

## Setup

Pay attention on the packages `sand` and `igraphdata`. The first one contains all the different network datasets used by [@kolaczyk2020statistical]. Then, the second one contain network data to be used together with the `igraph` package. More information can be found [here](https://cran.r-project.org/web/packages/sand/sand.pdf){target="_blank"} and [here](https://cran.r-project.org/web/packages/igraphdata/igraphdata.pdf){target="_blank"}. Feel free to explore these network data on your own time.

```{r setup, include=TRUE, results= 'hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(igraph)

library(sand)       
library(igraphdata)
```

## Aims of the practical

-   Simulate network models

-   Analyse their key attributes

-   Compare them

`igraph` enables us to generate networks based on some of the some key network models we discussed. I list below some of these functions:

-   `sample_gnp()`

-   `sample_smallworld()`

-   `sample_pa()`

Go to `igraph`'s [webpage](https://igraph.org/r/html/latest/){target="_blank"} and familiarise yourselves with the syntax of these functions.

```{block, type='alert alert-warning'}

**Task**: Create three networks using these three functions with, let's say, $n = 500$ nodes each.

Visualise and compare these networks.
```

Importantly, just before you generate a network with one of the above three functions you will need to define a random seed for reproducibility reasons, e.g. `set.seed(55)`.

```{block, type='alert alert-warning'}

**Question**: *Without* defining a random seed, generate a network as per the above, 
run some type of quick analysis about this network and then repeat the code by generating 
a second network with the exact same characteristics. Run the same analysis for the
second network. Are the results the same?

Why do we need to define the random seed ?
```

## Examples of modelled networks

```{r}

sample_gnp(n = 500, p = 0.02) %>% 
  plot(layout=layout_in_circle, vertex.label=NA)

sample_pa(n = 500, directed=FALSE) %>% 
  plot(layout=layout_in_circle, vertex.label=NA)

sample_smallworld(dim = 1, size = 500, nei = 5, p = 0.05) %>% 
  plot(layout=layout_in_circle, vertex.label=NA)

make_lattice(dim =1, length = 100, nei = 5) %>% 
  plot(vertex.label=NA)

```




```{block, type='alert alert-warning'}

**Challenge 1**: Think about the commuting datasets you used in the past tutorials. Is the commuters network Small World? Is it Scale-Free network?

**Challenge 2**: Can you demonstrate where would this graph sit on the scale between ranom and regular network using your knowledge centrality measures in I graph?

**Challenge 3**: Taking the commuters network or network of your choice, can you demonstrate the Bettencourt-West/Marshall’s law?

**Challange 4** Taking the data and model from the challange 3, can you find the value of $/beta$ (the exponent)? What does that value mean for your variables and the phenomena tehy represent?

```

Extra resources: 
- [Network Analysis in R from Dai Shizuka](https://dshizuka.github.io/networkanalysis/tutorials.html)
- [Statistical Analysis of Network Data with R](https://link.springer.com/book/10.1007/978-1-4939-0983-4)
- [Awesom Network Analysis: list of useful R packages (and much more)](https://github.com/briatte/awesome-network-analysis#r)
- [R Graph Gallery](https://r-graph-gallery.com/network.html)
_ [Static and dynamic network visualization with R from Katya Ognyanova](https://kateto.net/network-visualization)