---
title: "Accessibility in practice"
author: Emmanouil Tranos
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
editor: source
bibliography: references.bib
---

## Setup

To run this pratical you need to have an RStudio project as discussed in the 
[previous practical](https://etranos.info/urban_analytics_city_science/src/network_analysis.html). 

```{r setup, results= 'hide'}
library(tidyverse)
library(accessibility)
library(sf)
library(kableExtra)
library(rprojroot)
library(osrm)

path <- find_rstudio_root_file()

knitr::opts_chunk$set(include=TRUE, message=FALSE, warning=FALSE)
```

## Aims of the practical

The aim of this practical is to calculate different accessibility indicators.
For a brief overview about accessibility, please have a look at [@rodrigue2020geography]
(for example [here](https://transportgeography.org/contents/methods/transportation-accessibility/)
).
For a review have a look at @HOLL2007286, @el2022making and @shi2020literature
as well as at some older and well cited papers such as @geurs2004accessibility and 
@bruinsma1998accessibility. The below typology of accessibility concepts is from
@HOLL2007286.

1.  Network access

    -   distance to access the network

    -   travel opportunities

2.  Travel cost measures

    -   network access + distance/time travelled on the network

3.  Market potential accessibility

    -   destinations at greater distance provide diminishing opportunities

    -   $Acc_i = \sum_j\frac{W_j} {c_{ij}^a}$

Among other `R` packages, the practical is based on the `accessibility` package.
Some of the below materials are based on the package's [vignette](https://cran.r-project.org/web/packages/accessibility/vignettes/accessibility.html).

Let's start with using the toy data that comes with the `accessibility` package.
As you can see below, the data comes in `R`'s native data format called [`.Rds`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/readRDS)
and, therefore, we use the relevant function to read it -- `readRDS()`.

The first data object we are using is the travel time matrix between all potential
origins and destinations in city of Belo Horizonte, Brazil. 
As you can see, the matrix is stored in a *long format* and not as an $n*n$ matrix. 
If you check the dimension of that object you will understand what I mean
(`dim(travel_matrix)`).

```{r}
data_dir <- system.file("extdata", package = "accessibility")

travel_matrix <- readRDS(file.path(data_dir, "travel_matrix.rds")) 

head(travel_matrix)
```

Then, we are loading the characteristics of each location. As you can see, the 
`id` records match with the `from_id` and `to_id` from the `travel_matrix`.

```{r}
land_use_data <- readRDS(file.path(data_dir, "land_use_data.rds"))
head(land_use_data)
```

The first accessibility function we are using is `cost_to_closest()`. As per its
name, it calculates the minimum travel cost to the closest *n* number of opportunities.
In our example, as travel cost we understand the travel time between origins and 
destinations included in the column `travel_time` from the `travel_matrix` object. 
We are only interested in the closest (n=1) opportunity, which in this example is a school.

```{r}
mtc <- cost_to_closest(
  travel_matrix,
  land_use_data,
  opportunity = "schools",
  travel_cost = "travel_time",
  n = 1
)

head(mtc)
```

The next measure is the total number of opportunities (in other words cummulative
opportunities) within a cut-off of the travel cost. In other words, the below provides
the nu,ber of jobs accessibile within 30 minutes of travel time.

```{r}
cum_cutoff <- cumulative_cutoff(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  cutoff = 30
)

head(cum_cutoff)
```

If the above was a form of *active* accessibility -- that is how many opportunities
one can access from their location -- the below offers a *passive* form of accessibility
-- that is how many individuals can be reached from each destination within a cuttof.
The latter is based on `travel_time` in the below. 

```{r}
passive_cum_cutoff <- cumulative_cutoff(
  travel_matrix,
  land_use_data,
  opportunity = "population",
  travel_cost = "travel_time",
  cutoff = 30,
  active = FALSE
)

head(passive_cum_cutoff)
```

Last, but not least we can calculate the gravity based accessibility indicators,
or, in other words, potential accessibility indicators. This is the same measure 
as the example we discussed during the [class](https://etranos.info/urban_analytics_city_science/src/transport_geography.html#/spatial-interactions-and-the-gravity-model-2){target="_blank"}.
The intuition is that the furthest away each opportunity is, the less important
it becomes. So, the accessibility of a location $i$ is equal to the sum of all
opportunities available divided -- in other words discounted -- by the distance or 
the cost to reach them. The `negative_exp` function can take different decay functions
(see [here](https://cran.r-project.org/web/packages/accessibility/vignettes/decay_functions.html)).
We know from previous studies that the exponential function represents better shorter
interactions (e.g. within cities) while the power function works better for longer distance
interactions (e.g. migrations flows) [@fotheringham1989spatial, see also discussions
in @oshan2021spatial].

$Acc_{i} = \sum_j \frac{W_j}{d_{ij}^2}$


```{r}
negative_exp <- gravity(
  travel_matrix,
  land_use_data,
  opportunity = "schools",
  travel_cost = "travel_time",
  decay_function = decay_exponential(decay_value = 0.2)
)

head(negative_exp)
```

And now we can plot these metrics. The `grid` object is a toy spatial dataset of 
the places we calculate the above accessibility indicators. We first load these
spatial data, merge them with the acessibility indicator we want to plot and then
we plot these data.

```{r}
grid <- system.file("extdata/grid_bho.rds", package = "accessibility")
grid <- readRDS(grid)

spatial_data <- merge(grid, cum_cutoff, by = "id")

ggplot() +
  geom_sf(data = spatial_data, aes(fill = jobs), color = NA) +
  labs(
    title = "Job accessibility by transit in under 30 min.",
    fill = "Accessible jobs"
  ) +
  scale_fill_viridis_c() +
  theme_void()
```

## London example

All the above were based on the data that came together with the `accessibility`
package. Let's repeat this exercise using some real world data from [data.london.gov.uk](https://data.london.gov.uk/dataset/jobs-and-job-density-borough).
Jobs and population per Local Authority represent the opportunities. 

```{r}
jobs <- read_csv("https://data.london.gov.uk/download/jobs-and-job-density-borough/79e85749-e38d-4bb6-8327-c60c6be27222/Jobs_and_Job_Density.csv") %>% 
  filter(year == 2020) %>%  # I am only using 2020 data. Remove this line 
                            # to see what will happen to the data.
  glimpse()

pop <- read_csv("https://data.london.gov.uk/download/land-area-and-population-density-ward-and-borough/77e9257d-ad9d-47aa-aeed-59a00741f301/housing-density-borough.csv") %>% 
  filter(Year == 2020) %>% # This is a projection, but it will work for now
  glimpse()
```

Jobs and pop include observations for spatial units other than London's 33 Local
Authorities. Explore `pop` and `jobs` to see that.
Therefore, we use the `inner_join()` function to join them. See 
[here](https://dplyr.tidyverse.org/reference/mutate-joins.html) how it works. 
You should be able to understand the different functions to join data.

```{r}
df <- inner_join(jobs, pop, by = c("code" = "Code")) %>% 
  filter(area != "London") %>% 
  rename(id = code) 
```

To produce a nicer looking table I am using here the `kableExtra` package.
It just adds some nice styling for `html` outputs.

```{r}
df %>% kbl() %>%
  kable_styling(full_width = F) %>%   #Nice(r) table
  scroll_box(width = "500px", height = "200px")
```

Next, we are bringing the spatial data -- that is the polygons -- for London's Local
Authorities. We start by downloading all the polygons for all Local Authorities 
in the country. Please go to this [webpage](https://geoportal.statistics.gov.uk/datasets/1119a90ec5f343678f044374392e6bda_0/explore), download the `geojson` file and save it to the appropriate `data` folder 
within your project directory.

However, we don't need all the Local Authorities, but instead we only need to ones
within London. Therefore we use the `filter` option to select only the Local Authorities
which are included within -- this is the `%in%` operator -- the `df$id` column. 
Finally, we create a new object which only includes the centroids of these polygos?
Why are we doing this? Because we need to calculate distances and accessibilities
metrics for specific points and we will use these points as representative for each
Local Authority. 

::: callout-tip
Do you see any problems with this approach?
:::

```{r}
path.shape <- paste0(path, "/data/Local_Authority_Districts_(May_2021)_UK_BFE.geojson")

london <- st_read(path.shape) %>%
  dplyr::filter(LAD21CD %in% (df$id))

london.centres <- st_centroid(london)
```

Now we need to actually calculate the distance matrix. We are going to use a real
world road network to perform what is known as *routing*. There is a great discussion
about routing and before you move forward I would encourage you to read the relevant
section from @lovelace2019geocomputation, which can be found [here](https://geocompr.robinlovelace.net/transport.html#routes).

We are using the `osmr` package which can do routing for us. The details of the `osrm`
package can be founf [here](https://github.com/riatelab/osrm).

The below code creates a distance table between all the `london.centres`. We are 
interested in distance measured as time -- i.e. duration -- by car. The `osrm`
package offers other options too.

```{r}
dist <- osrmTable(loc = london.centres, 
                  measure = "duration",
                  osrm.profile = "car")
```
The output of the above is a list with an $n*n$ distance matrix, and a dataframe
with the longitude and latitude of the origins and destinations. We are mostly
interested in the former, so we start the data wrangling by creating a `help` 
dataframe with these distances, or in other words, durations.
We then add row and column names using the `london.centres` names. We can just
do that because the matrix follows the same order as the intital spatial data
of Local Authorities. I always like to throw in a `dim` function just to keep 
an eye on the dimensions of the objects I create. Indeed, the `dist` object is 
a $33*33$ matrix as it should have been.

Then, I use the function `odmatrix_to_od()` form the `od` package. Why? Because
we need to convert the $n*n$ matrix to a long format. If you remember, we used to
call this format an [edge list](https://etranos.info/urban_analytics_city_science/src/intro_networks.html#/directed-weighted-networks-1).

And at the end
```{r}
help <- dist$durations 


rownames(help) <- as.vector(london.centres$LAD21CD)
colnames(help) <- as.vector(london.centres$LAD21CD)

dim(dist$durations)

# od::odmatrix_to_od(dist$durations)

dist.df <- as.data.frame(as.table(help)) %>% 
  rename(from_id = Var1,
         to_id = Var2,
         travel_time = Freq)
```
As before, we calculate the cumulative opportunities for London's Local Authorities
(the centroids to be more accurate) within a cutoff of 20 minutes.

```{r}
mtc <- cumulative_cutoff(
  dist.df,
  df,
  opportunity = "number_of_jobs",
  travel_cost = "travel_time",
  cutoff = 20
)
head(mtc)
```

And finally, we plot our accessibility map.

```{r}

london <- merge(london, mtc, by.x = "LAD21CD", by.y = "id" )
  
  
ggplot() +
  geom_sf(data = london, aes(fill = number_of_jobs), color = NA) +
  labs(
    title = "Job accessibility by driving in under 20 min.",
    fill = "Accessible jobs"
  ) +
  scale_fill_viridis_c() +
  theme_void()
```

::: callout-tip
Can you calculate and plot other accessibility metrics for London?
:::

::: callout-tip
What do we miss by focusing on London' accessibility?
:::


