---
title: "P1: APIs and urban data"
author: Emmanouil Tranos
date: "`r Sys.Date()`"
editor: source
output:
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

## Aims of the practical

-   Understand the technicalities and the breadth of the urban data landscape

-   Understand the *messiness* of real world (urban) data

-   Access programmatically urban data and understand the benefits of such data access practise.

## APIs

Application Programming Interface (APIs) represent a 'gate' or otherwise a platform that enables a client (that is you) to interact with a server (for example [opendata.bristol.gov.uk](https://opendata.bristol.gov.uk/)). According to @amazonWhatAPI:

> In the context of APIs, the word Application refers to any software with a distinct function. Interface can be thought of as a contract of service between two applications. This contract defines how the two communicate with each other using requests and responses. Their API documentation contains information on how developers are to structure those requests and responses.

The client's software (this might be R for example) sends a request to the server requesting specific data. The response is the data the client asked.

More commonly, the client might be a mobile phone app (e.g. train network status app) and the server is the network operator's server.

APIs can be private or public types. For more inthe description from @amazonWhatAPI [here](https://aws.amazon.com/what-is/api/#:~:text=API%20stands%20for%20Application%20Programming,other%20using%20requests%20and%20responses.)

## JSON file format

Although some urban data can be downloaded as a tidy[^1] table in a `.csv` format, most APIs will ship data to you in a `JSON` format.

[^1]: Every column is a variable; every row is an observation; every cell is a single value. See \[here\])(https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) for more info

According to the official JSON website [@jsonJSON], JSON

> is a lightweight data-interchange format. It is easy for humans to read and write. It is easy for machines to parse and generate. It is based on a subset of the JavaScript Programming Language Standard ECMA-262 3rd Edition - December 1999. JSON is a text format that is completely language independent but uses conventions that are familiar to programmers of the C-family of languages, including C, C++, C#, Java, JavaScript, Perl, Python, and many others. These properties make JSON an ideal data-interchange language.

It adopts a hierarchical structure such as `attribute:value` instead of the fixed structure of tidy data. Following @singleton2017urban, such `attribute:value` pairs might be:

-   "mean income":"50000"

-   "internet speed":"35"

-   "jobs":"15000"

These attributes might be organised hierarchically:

("city":"Bristol,"mean income": "27000", "internet speed": "55", "jobs":("2019":"15000", "2020":"14000")), ("city":"Newcastle,"mean income": "25000", "internet speed": "45", "jobs":("2019":"11000", "2020":"10000"))

Although this format is more verbose (attribute names are repeated) it is more flexible (not all observations need to have all attributes) [@singleton2017urban]

## Practical example with API

[opendata.bristol.gov.uk](https://opendata.bristol.gov.uk/pages/homepage/) offers a lot of data about Bristol. For this example, let's try to work with [Jobs by Ward (by industry) in Bristol 2020](https://opendata.bristol.gov.uk/datasets/bcc::jobs-by-ward-by-industry-in-bristol-2020/explore?location=51.470876%2C-2.616185%2C12.13). As you can see, this is a small data set with $34$ observations representing Bristol's $34$ Wards. You can visualise these data with the `Table` tab on the right hand-side and even download a `.csv` with these data. To begin with, do both of these things.

Then, go to the `More Info` icon, scroll down on the left hand side and click on `I want to use this`. You can find the relevant link under `View API Resources`. We are interested in the `GeoJSON`. You can view the data in this format and also copy the links to access these data *programmatically*.  The *weird* output on your screen is the data you requested in `JSON` format. If you carefully read the URL from right to left you can see some logic on how this URL is built.

```{block, type='alert alert-warning'}

**Question**: Why should you go through all these efforts to build an API request and 
download the data in `JSON` file format instead of just downloading a `.csv` file?

```

So, let's try to access these data programmatically. First, we need to load the relevant `R` packages

```{r setup, message=FALSE, results='hide', error=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
```

To access the data, we are going to use the function `fromJSON()` from the `jsonlite` package. Please check the syntax of this function as well as the other function of this package.

```{r get data}
dat <- jsonlite::fromJSON("https://services2.arcgis.com/a4vR8lmmksFixzmB/ArcGIS/rest/services/Jobs_by_Ward_(by_industry)_in_Bristol_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", flatten = TRUE)
```

We now have an `R` object called `dat`. Let's try to see how it looks like

```{r type}
typeof(dat)
```

So, this is a list. If you need a reminder for what a list is, see [here](http://adv-r.had.co.nz/Data-structures.html). Now the difficult part. Let's try to figure out the structure of this `JSON` formatted data.

```{r structure}
str(dat)
```

Well, it's complicated. Looking on the above output of `list()` the object that looks most useful to us is the `dat$features`.

```{block, type='alert alert-warning'}

**Question**: Based on the output of `list()`, why do you think is `dat$features` 
the most useful part of `dat` for us?

```

And if you want to see how the dataframe `dat$features` looks like, the below function will show the first five rows.

```{r head}
glimpse(dat$features)
```

Looking into the above, I can think of the the following necessary data cleaning and wrangling steps:

1.  remove the first five columns as they don't include any useful data

2.  remove the string "properties." from the column names

3.  although not necessary, it is common practice for the first columns to contain identification data such as the ward names and codes.

To do all the above, we will use the below functions from the `dplyr` package, which is part of the [`tidyverse` ecosystem](https://www.tidyverse.org/).

```{r wrangling}
dat <- dat$features %>% 
  select(-type, -id, -geometry.type, -geometry.coordinates, -properties.OBJECTID, -properties.Shape__Area, 
         -properties.Shape__Length) %>%                       # remove useless columns
  rename_with(~sub("properties.", "", .)) %>%                 # ~ introduces a formula: function `sub` substitutes (in other words replaces) the string "properties." with an empty sting "" for all columns (symbol .) 
  relocate(Ward, Ward_code_1) %>%                             # looks better
  rename(Ward_code = Ward_code_1) %>%                         # as above, Ward_code looks better than Ward_code_1
  glimpse()                                                   # a glimpse of the output
```

Let's try now to do some analysis based on these data and create some outputs.

```{r analysis}
dat %>% mutate(total = rowSums(across(c(4:21))),           # total jobs in a ward
               across(c(4:21),                             # select the columns you are interested in
                      .names = "{col}_share") / total) %>% # divide these columns by total to create industry share within ward  
                                                           # and name the new columns as oldname_share
  select(1:2, 23:40) %>%
  pivot_longer(!c(Ward, Ward_code),                        # These are the columns I am 'pivoting'
               names_to = "industry",                      # This is how I will name the column with all the industries ...
               values_to = "jobs_share") %>%               # ... and the column with the jobs percentage
                                                           # Look https://tidyr.tidyverse.org/articles/pivot.html to 
                                                           # understand the long to wide / wide to long transformations.
  filter(Ward == "Central") %>%                            # Let's focus on Central Bristol
  ggplot(aes(x=industry, y=jobs_share)) +                  # ... and create a bar chart for it jobs distribution
    geom_bar(stat = "identity") +
# to understand the "identity part go on the Details section of the geom_bar documentation:
# https://www.rdocumentation.org/packages/ggplot2/versions/1.0.1/topics/geom_bar
    
    scale_x_discrete(guide = guide_axis(angle = 90)) +     # This line rotates the x labels,
    ylab("jobs %")                                         # and here we rename the y axis. 
```

Now, let's do something slightly more interesting. Let's create two plots to compare the job distribution between two Bristol wards.

```{r comparison}
dat %>% mutate(total = rowSums(across(c(4:21))),           # total jobs in a ward
               across(c(4:21),                             # select the columns you are interested in
                      .names = "{col}_share") / total) %>% # divide these columns by total to create industry share within ward  
                                                           # and name the new columns as oldname_share
  select(1:2, 23:40) %>%
  pivot_longer(!c(Ward, Ward_code),                        # These are the columns I am 'pivoting'
               names_to = "industry",                      # This is how I will name the column with all the industries ...
               values_to = "jobs_share") %>%               # ... and the column with the jobs percentage
                                                           # Look https://tidyr.tidyverse.org/articles/pivot.html to 
                                                           # understand the long to wide / wide to long transformations.
  filter(Ward == "Central" |
         Ward == "Easton") %>%                             # Let's focus on Easton and Central Bristol
  ggplot(aes(x=industry, y=jobs_share)) +                  # ... and create a bar chart for it jobs distribution
    geom_bar(stat = "identity") +
    scale_x_discrete(guide = guide_axis(angle = 90)) +     # This line rotates the x labels,
    ylab("jobs %") +                                       # and here we rename the y axis. 
    facet_wrap(vars(Ward), ncol = 2)                       # facet_wrap creates two graphs based on `ward` and puts one next to each other
```

Well done! You've just utilised Bristol's open data and became part of the urban open data revolution 😃.

Obviously the `x` labels are rather ugly.

```{block, type='alert alert-warning'}
**Task:** using and tweaking the code provided here, can you rename these labels? 
```

```{block, type='alert alert-warning'}
**Next task:** Search for some other interesting data from 
[opendata.bristol.gov.uk](https://opendata.bristol.gov.uk/),
access it programmatically, do the necessary data wrangling and produce some 
graphs in a new `.Rmd` document. Make sure to knit it as `.html`.

If this is to easy for you, why don't you check the new [Quarto format](https://quarto.org/docs/guide/).
```

## Accessing APIs using existing `R` packages

Now you might be lucky and somebody else has already done all the infrastructural / back-end work in order for you to easily interrogate an API and access their data. How would this look like? Like another `R` package. Here is a list with a few examples:

-   `nomisr` is a package that allows you to access [Nomis](https://www.nomisweb.co.uk/) data with `R`. Find [here](https://cran.r-project.org/web/packages/nomisr/vignettes/introduction.html) an introduction.

-   `tidysencus` is part of the `tidyverse` ecosystem and offers access to the US Census. This is the [link](https://walker-data.com/tidycensus/). Pay extra attention to the *spatial data* option.

-   The city of Toronto offers an `R` [package](https://sharlagelfand.github.io/opendatatoronto/) to access [City of Toronto Open Data Portal](https://open.toronto.ca/).

-   See a relevant book by @apis available [here](https://bookdown.org/paul/apis_for_social_scientists/).

```{block, type='alert alert-warning'}
**Another task:** On your own time, play with these packages and access some of the 
(urban) data they offer.
```

## References
